Density-dependent selection and the limits of relative fitness

### 2017-12-16 manuscript.tex
l

Evolutionary biologists frequently describe selection using relative fitness, and relative fitness models are foundational in population genetics. Yet the population ecological basis of relative fitness is poorly understood, and the classical ecology literature on selection in crowded populations seems to be incompatible with widely-used relative fitness models such as the Wright-Fisher. Here we develop a generalization of the Wright-Fisher model in which density depends dynamically on the demographic rates of the types present. We then explore the population ecology of relative fitness using this model as a reference point. Although density-dependent selection can cause relative fitness to break down, relative fitness models are fairly robust in populations close to demographic equilibrium. In particular, relative fitness only breaks down if selection is strong and density-dependent, and density is also selection-dependent. Our generalized Wright-Fisher model clearly distinguishes between three demographic parameters, each of whose behavior is, on its own, correctly described by relative fitness models. In contrast, the classical literature on density-dependent selection confounds them. We argue that selection-independent density is ecologically plausible for many species given the prevalence of reproductive excesses, an important aspect of selection omitted in many ecological models. Our model also offers a natural alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilibrium.

@@

@@

Equipped with our new model, we turn to the evaluation of Eq.~\eqref{eq:canonical}. We first discuss selection on the ability to contest territories, which behaves like a pure constant-$N$, relative fitness trait, and discuss how this fits with MacArthur's analysis of selection in crowded populations (section ``$K$-selection and selection-dependent density''). We then discuss selection on density-regulating traits, contrasting the results from the classical density-dependent selection literature with our model (section ``Density-regulating traits and the threat of strong selection''). We conclude [...]

@@

Summarizing the behaviour of the three traits in the variable-density lottery: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent. Despite their differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the canonical selection equation. This rich variety of behaviors in relation to density is quite different from that found in the classical density-dependent selection literature \citep{roughgarden_1979,christiansen_2004}.

To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme posited a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater saturation density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Thus, setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in that they are both density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep). Thus, sufficiently strong selection is all that is needed for relative fitness to fall apart in the classical density-dependent selection literature.

Our variable-density lottery model shows that this contrast is not simply a lack of ecological realism on the part of relative fitness models. Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation). As noted in the introduction, this precludes the important possibility that selection is concentrated at a fragile juvenile stage as a result of a  reproductive excess \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.

In our variable-density lottery, reproductive excess appears when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition. In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types can grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection, artificially invalidating Eq.~\eqref{eq:canonical}. In this respect, the Wright-Fisher model and similar viability selection heuristics actually capture an important ecological process.

We now turn to the breakdown of Eq.~\eqref{eq:canonical}. We first discuss the problem shown in Fig.~\ref{fig:strengthofselection}, which occurs when strong selection changes (adult) population density and is also density-dependent. In the variable-density lottery, this occurs if and only if types differ in more than one trait. The $c$ and $d$ traits highlight the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}. The combination is needed to pose a threat to Eq.~\eqref{eq:canonical}, and even then, the $b$ trait remarkably demonstrates that both may not be sufficient (recall that we are still assuming $N$ is not far from equilibrium). In this respect the simple linear models that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012} actually represent a complicated form of the density-selection interaction that confounds a key underlying issue.

@@

Relative fitness models truly break down when $N$ is far from equilibrium and selection is density-dependent. For example, wild \textit{Drosophila} experience huge seasonal boom-bust cycles in population density coupled to strong selection that drives large swings in allele frequency \citep{bergland_14}. In this case there is no choice but to abandon relative fitness, and our model provides one potentially suitable option. Whether or not our density-dependent lottery model is a good description of \textit{Drosophila} ecology, the close connection between our model and Wright-Fisher is useful, because drift in our model should behave broadly similarly. Thus, our model it should provide a useful starting point for analyzing evolution in this and other from-from-equilibrium situations.

Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. A variety of approaches have been developed for dealing with these issues in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}. Although this manuscript, and density-dependent selection more broadly, are pertinent to the interplay between ecology and evolution \citep{travis_2013}, our focus here is the description of the time-dependent process by which selection changes allele frequencies. This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.

@@

In section ``Density-dependence and the strength of selection'' we argued that the density-dependent factor $f(\overline{b},N)$ is unchanged at the beginning and end points of an equilibrium-to-equilibrium $b$. Here we estimate the magnitude of the deviation in $f(\overline{b},N)$ during the sweep.

>>>>

Relative fitness  models are the basis of  foundational in population genetics.

Yet the population ecological basis of relative fitness is poorly understood, and the classical ecology literature on selection in crowded populations seems to be incompatible with widely-used relative fitness models such as the Wright-Fisher. Here we develop a generalization of the Wright-Fisher model in which density depends dynamically on the demographic rates of the types present. We then explore the population ecology of relative fitness using this model as a reference point. Although density-dependent selection can cause relative fitness to break down, relative fitness models are fairly robust in populations close to demographic equilibrium. In particular, relative fitness only breaks down if selection is strong and density-dependent, and density is also selection-dependent. Our generalized Wright-Fisher model clearly distinguishes between three demographic parameters, each of whose behavior is, on its own, correctly described by relative fitness models. In contrast, the classical literature on density-dependent selection confounds them. We argue that selection-independent density is ecologically plausible for many species given the prevalence of reproductive excesses, an important aspect of selection omitted in many ecological models. Our model also offers a natural alternative to relative fitness when the latter is untenable, as is likely the case far from demographic equilibrium.

@@

Another issue with the constant-$N$ relative fitness description of selection is that it precludes consideration of longer-term aspects of the interplay between evolution and ecology such as population extinction. A variety of approaches have been developed for dealing with these issues in quantitative genetics \citep{burger1995evolution,engen_2013}, population genetics \citep{bertram2017predicting} and adaptive dynamics \citep{ferriere2013eco,dieckmann2004adaptive}. Although density-dependent selection is  pertinent to these longer-term issues \citep{travis_2013}, our focus here is the description of the time-dependent process by which selection changes allele frequencies. This is particularly critical for making sense of evolution at the genetic level, for which we now have abundant data.

@@

Equipped with this new model, we turn to the evaluation of Eq.~\eqref{eq:canonical}. We first discuss selection on the ability to contest territories, which behaves like a pure constant-$N$, relative fitness trait, and discuss how this fits with MacArthur's analysis of selection in crowded populations (section ``$K$-selection and selection-dependent density''). We then consider selection on density-regulating traits (section ``Density-regulating traits and the threat of strong selection''), and conclude by contrasting the classical density-dependent selection literature with our results (``Discussion'').

@@

Summarizing the three traits in the variable-density lottery model: (i) $c$-selection is density-dependent, but $c$ does not regulate density; (ii) $d$ regulates density, but $d$-selection is density-independent; (iii) $b$ regulates density and $b$-selection is density-dependent. Despite these differences, pure $b$, $c$ and $d$ sweeps starting and ending at equilibrium all obey the canonical selection equation. This rich variety of behaviors in relation to density is quite different from that found in the classical density-dependent selection literature \citep{roughgarden_1979,christiansen_2004}.

To briefly review: based on a diploid, bi-allelic variant of the logistic model, the $r$/$K$ scheme proposed a dichotomy between $r$-selection (uncrowded) and $K$-selection (crowded) \citep{macarthur_1962}, with the latter taken to mean selection for greater saturation density \citep{gill_1974}. A more general Lotka-Volterra model introduces the inter-type $\alpha_{ij}$ competition coefficients, with selection on these termed ``$\alpha$-selection'' \citep{gill_1974,joshi_2001}. Setting aside $r$ which confers no selective advantage at equilibrium, we are left with $K$ and $\alpha$, which both behave like $\delta$ in that they are density-dependent and cause density to change over a sweep (although $N$ only dips temporarily during an $\alpha$-sweep). Thus, strong selection is sufficient for relative fitness to break down in the classical view of density-dependent selection. Indeed, in the defense of Eq.~\eqref{eq:canonical} given by \cite{kimura1969natural}, it was necessary to assume that $s$ will be a few percent at most.

Our variable-density lottery model shows that it is not simply a lack of ecological realism that underlies the incompatibility of relative fitness with the classical view of density-dependent selection. Rather, in many population growth models, only one life-history stage is represented, and the competitive effects resulting from crowding appear as a reduction in absolute fitness that only depends on the type densities at this life-history stage (e.g. the $n_i^2$ and $n_in_j$ terms in the Lotka Volterra equation). As noted in the introduction, this precludes selection concentrated at a fragile juvenile stage as a result of a reproductive excess \citep{chesson_1983,turner1968population,kimura1969natural,nei1971fertility}.

Reproductive excesses appear in the variable-density lottery model when the number of propagules is greater than the number of available territories. Then only $\approx 1/L$ of the juveniles contesting available territories survive to adulthood. Unlike the role of adult density $n_i$ in single-life-stage models, it is the propagule densities $l_i$ that represent the crowding that drives competition. In general, reproductive excesses will tend to produce strictly-relative lottery-type contests in which fitter types grow at the expense of others by preferentially filling the available adult ``slots''. The number of slots can remain fixed or change independently of selection at the juvenile stage. By ignoring reproductive excesses, single life-stage models are biased to have total population density be sensitive to ongoing selection. In this respect, the Wright-Fisher model and similar viability selection heuristics actually capture an important ecological process.

We now turn to the breakdown of Eq.~\eqref{eq:canonical}. We first discuss the problem shown in Fig.~\ref{fig:strengthofselection}, which occurs when strong selection changes population density and is also density-dependent. In the variable-density lottery, this occurs if and only if types differ in more than one trait. The $c$ and $d$ traits represent the two distinct directions in which density and selection interact: selection may depend on density, and density may depend on selection \citep{prout_1980}. The combination is necessary to pose a threat to Eq.~\eqref{eq:canonical}. However, the $b$ trait remarkably demonstrates that the combination is not sufficient, since the density-dependence of $b$-selection disappears over equilibrium-to-equilibrium $b$-sweeps. Thus, the simple linear models that have become standard in discussions of density-dependent selection \citep{roughgarden_1979,christiansen_2004,mallet_2012,travis_2013} actually represent a complicated form of the interaction between density and selection, and their parameters confound the underlying issues.

@@

Relative fitness models truly break down when $N$ is far from equilibrium and selection is density-dependent (as seems likely \citep{travis_2013}). For example, wild \textit{Drosophila} experience large seasonal boom-bust cycles in population density coupled to strong selection that drives large swings in allele frequency \citep{bergland_14}. In this case there is no choice but to abandon relative fitness, and our model provides one potentially suitable option. Whether or not our density-dependent lottery model is a good description of \textit{Drosophila} ecology, the close connection between our model and Wright-Fisher is useful, because drift in our model should behave broadly similarly. Thus, our model it should provide a useful starting point for analyzing evolution in this and other far-from-equilibrium situations.

@@

In section ``Density-regulating traits and the threat of strong selection'' we argued that the density-dependent factor $f(\overline{b},N)$ is unchanged at the beginning and end points of an equilibrium-to-equilibrium $b$. Here we estimate the magnitude of the deviation in $f(\overline{b},N)$ during the sweep.
